\documentclass{article}
\usepackage{amsmath}
\begin{document}

\title{Proof: The Square Root of 2 is Irrational}
\author{}
\date{}
\maketitle

\section*{Problem}
Prove that $\sqrt{2}$ is an irrational number.

\section*{Solution 1: Proof by Contradiction}
Assume, for contradiction, that $\sqrt{2}$ is rational. This means it can be written as a fraction of two integers in lowest terms:
\begin{equation}
    \sqrt{2} = \frac{p}{q}, \quad \text{where } p, q \text{ are integers with } \gcd(p, q) = 1.
\end{equation}

Squaring both sides:
\begin{equation}
    2 = \frac{p^2}{q^2}.
\end{equation}

Multiplying both sides by $q^2$:
\begin{equation}
    2q^2 = p^2.
\end{equation}

Since $p^2$ is divisible by $2$, it follows that $p$ must also be divisible by $2$ (as squares of odd numbers are odd). So we write:
\begin{equation}
    p = 2k \quad \text{for some integer } k.
\end{equation}

Substituting into the equation:
\begin{equation}
    2q^2 = (2k)^2 = 4k^2.
\end{equation}

Dividing both sides by $2$:
\begin{equation}
    q^2 = 2k^2.
\end{equation}

This implies that $q^2$ is also divisible by $2$, so $q$ must also be divisible by $2$. This contradicts our original assumption that $p$ and $q$ have no common factors. Therefore, $\sqrt{2}$ is irrational.

\section*{Solution 2: Proof Using Prime Factorization}
Suppose $\sqrt{2}$ were rational, meaning $\sqrt{2} = \frac{p}{q}$ for some integers $p$ and $q$ with no common factors. Squaring both sides gives:
\begin{equation}
    p^2 = 2q^2.
\end{equation}

In prime factorization, $p^2$ has an even number of factors, and $2q^2$ has an odd number of factors of $2$. This contradiction shows that $p^2$ cannot equal $2q^2$, proving that $\sqrt{2}$ is irrational.

\section*{Solution 3: Proof Using Infinite Descent}
Suppose $\sqrt{2}$ is rational, so $\sqrt{2} = \frac{p}{q}$ in lowest terms. Then:
\begin{equation}
    p^2 = 2q^2.
\end{equation}

Since $p^2$ is even, $p$ is even, so we write $p = 2k$. Substituting:
\begin{equation}
    (2k)^2 = 2q^2 \Rightarrow 4k^2 = 2q^2 \Rightarrow q^2 = 2k^2.
\end{equation}

Thus, $q^2$ is even, meaning $q$ is also even. This creates an infinite descent of smaller and smaller even numbers, contradicting the assumption that $p$ and $q$ are in lowest terms. Hence, $\sqrt{2}$ is irrational.

\end{document}
