\documentclass{article}
\usepackage{amsmath}
\begin{document}

\title{Proof: The Sum of Two Even Integers is Even}
\author{}
\date{}
\maketitle

\section*{Problem}
Prove that the sum of two even integers is always even.

\section*{Solution 1: Direct Proof Using Definition}
Let $a$ and $b$ be two even integers. By definition, an even integer can be written as:
\begin{equation}
    a = 2m, \quad b = 2n, \quad \text{where } m, n \text{ are integers}.
\end{equation}

The sum of $a$ and $b$ is:
\begin{equation}
    a + b = 2m + 2n.
\end{equation}

Factoring out $2$:
\begin{equation}
    a + b = 2(m + n).
\end{equation}

Since $m + n$ is an integer, we conclude that $a + b$ is even.

\section*{Solution 2: Proof by Contradiction}
Assume, for contradiction, that the sum of two even integers is odd. That is, assume:
\begin{equation}
    a + b = 2k + 1, \quad \text{for some integer } k.
\end{equation}

Since $a$ and $b$ are even, they can be written as:
\begin{equation}
    a = 2m, \quad b = 2n.
\end{equation}

Adding these values:
\begin{equation}
    a + b = 2m + 2n = 2(m+n).
\end{equation}

Since $2(m+n)$ is even, this contradicts our assumption that $a + b$ is odd. Therefore, the sum of two even numbers must be even.

\section*{Solution 3: Proof Using Parity}
An integer is even if it has remainder $0$ when divided by $2$. That is,
\begin{equation}
    a \equiv 0 \pmod{2}, \quad b \equiv 0 \pmod{2}.
\end{equation}

Adding both congruences:
\begin{equation}
    a + b \equiv 0 + 0 \equiv 0 \pmod{2}.
\end{equation}

Since $a + b \equiv 0 \pmod{2}$, it follows that $a + b$ is even.

\end{document}
